\textsc{Rils}-\textsc{Rols} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-14 paper_resources/doc/paper.tex
Update...

	\author[1]{Aleksandar Kartelj}
	\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
	\address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
		In this paper, we solve the well-known, prominent  Symbolic regression problem, intensively studied within the last two decades. This problem has a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards establishing physical theory behind, thus helping to establish reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modelling analytic representations of the exciton binding energy in energy science,   finding models for band gaps of NaCl-type compounds
		in material science, just to name a few. We propose a new meta-heuristic-based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are several key concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used in the search to make it efficient with  determining the best--fitting   coefficients in considered solutions (models);  ($iii$) a novel fitness function is utilized which combines the three known measures: $R^2$ score, RMSE score, and the size of the model which is weighted by a parameter. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} algorithm to 14 other competitors from the literature where different levels of the Gaussian white noise in the input data are utilized.  Our method obtains new best exact percentages on both benchmark sets compared to other literature approaches regardless of varying the level of nose. Moreover, a new unbiased randomly generated set of problem instances are introduced where the scalability of our method is thoroughly demonstrated in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

	\begin{enumerate}
		\item The proposed method outperforms all (14) state-of-the-art methods for ground-truth problems from literature, considered in the SRBench benchmark.

		\item The method shows its high robustness, which is proved by comparison to other algorithms under different levels of Gaussian white noise.

		\item The method is very efficient, taking in average around xyz seconds to reach exact solution -- when exact solution is found. \fxnote{Ovo treba dodati u eksp. sekciji...}

		\item New set of unbiased instances is   introduced that are randomly generated formula of various sizes and number of input variables in exact models. This set was employed to methodologically analyze the influence of formula size, number of input variables and noise on the solving difficulty. 
	\end{enumerate}

@@

In order to check the statistical significance of the obtained exact percentages and their differences between the competitor approaches, we employed the statistical methodology comparing the results of \textsc{Rils}--\textsc{Rols} to 14 other approaches on all problem instances from the both ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz}. Note that an indicator random variable is involved per each run of the algorithms. Thus, the score that comes into the statistical observation is an indicator (0 or 1) indicating weather or not an algorithm was able to return the exact model within  considered run. Thus, we involved 1300 results into our statistical evaluation per each algorithm. Note that at some cases some of the 14 competitors are not known from \url{https://cavalab.org/srbench/results/} due to some reasons. For these cases, we simply count a 0 as the score.

Initially, Friedman’s test was separately executed for all competitor approaches.      In those cases in which the null hypothesis $H_0$ was rejected ($H_0$ states that there are no statistical differences between the obtained results of all competitors) pairwise comparisons are further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots. In each CD plot, the (15) competitor approaches are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is drawn.   The corresponding three (one for each level of noise) CD plots are shown in Figure~\ref{fig:cd-plots-exact-pcts}. The following conclusions may be drawn from there.

\begin{itemize}
	\item  Concerning the instance problems where no noise is included, \textsc{Rils}-\textsc{Rols} method returns the best average ranking in terms of obtained exact models.  The second best average ranking is achieved by \textsc{AI-Feynman} algorithm. There is  statistically significant difference between these two approaches w.r.t.\ number of found exact models. The third best approach is \textsc{Apf-Fe},  which is far behind \textsc{Rils}-\textsc{Rols} w.r.t.\ delivered average raking.  
	\item    Concerning the instance problems with the  level of noise equal to 0.001, \textsc{Rils}-\textsc{Rols} method again obtains the best average ranking in terms of the found exact models.   The second best ranking is obtained by \textsc{AI-Feynman} algorithm. There is again a statistically significant difference between the results obtained by these two approaches. The third best approach is \textsc{Afp-Fe} which perform statistically worse from \textsc{AI-Feynman}.  
	\item  Concerning the instance problems with the noise level of  0.01, again, \textsc{Rils-Rols} method produces the best average ranking according to obtained exact models; the second best approach w.r.t.\ average ranking is \textsc{Afp-Fe}, followed by \text{Dsr}, and \textsc{Afp}. These latter three approaches perform statistically equal, but significantly worse than \text{Rils}-\textsc{Rols} method. 	The fourth best approach according to its average ranking is \textsc{AI-Feynman} which is on pair with \textsc{gplearn}. Note that the average ranking of \textsc{AI-Feynman} method declines by increasing  the level of noise in the input data. Thus, despite being efficient on no-noisy data, it seem that \textsc{AI-Feynman} is very sensitive to the noise, which is not that dramatically conspicuous with our \textsc{Rils}-\textsc{Rols} method.

	 \fxnote{Marko: update text}
\end{itemize}

>>>>

	\author[1]{Aleksandar Kartelj}
	\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
	\address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
		In this paper, we solve the well-known, prominent  Symbolic regression problem, intensively studied in the literature within the last two decades. This problem has a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards establishing physical theory behind, thus helping to establish reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modelling analytic representations of the exciton binding energy in energy science,   finding models for band gaps of NaCl-type compounds
		in material science, just to name a few. We propose a new meta-heuristic-based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve this problem. There are several key concepts utilized in this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used in the search to make it efficient by quickly determining the best--fitting   coefficients in considered solutions (models);  ($iii$) a novel fitness function is utilized which combines the three known measures: $R^2$ score, RMSE score, and the size of the model which is weighted by an internal parameter. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} algorithm to 14 other competitors from the literature with the presence of different levels of the Gaussian white noise in the input data.  Our method is able to obtain new best exact percentages on both benchmark sets compared to other literature approaches regardless of varying the level of nose. Moreover, the performance of \textsc{Rils}-\textsc{Rols} is less sensitive with the increase of the level of noise in the data, as it is the other approaches. 
		Additionally, a new unbiased randomly generated set of problem instances are introduced where the scalability of our method is thoroughly demonstrated in terms of various instance properties. \textsc{Rils}-\textsc{Rols} method can be considered as a new state-of-the-art method for solving the problem of Symbolic regression. %for the non blck box problem instances?
	\end{abstract}
	\maketitle

@@

	\begin{enumerate}
		\item The proposed method outperforms 14 state-of-the-art methods for the problem instances from the two ground-truth benchmark sets from literature, used in the benchmarking platform SRBench.

		\item The method shows its high robustness, which is proved by its comparison to the other algorithms in presence of different levels of Gaussian white noise in the input data.

		\item The method is very efficient, taking in average less than 3 minutes to reach the exact solution, on the problem instances where it was possible. For example, the second best approach in terms of the delivered exact percentage, \textsc{AI-Feynman}, needs significantly more time, which is around 45 minutes.

		\item A new set of unbiased instances is introduced that consists of randomly generated formulae of various sizes and number of input variables presenting exact models that need to be found. This set was employed to analyze the effect of the model size, the number of input variables in the model and the level of noise on the solving difficulty. 
	\end{enumerate}

@@

In order to check the statistical significance of the results w.r.t.\ obtained exact models and their differences between the competitor approaches, we employed the statistical methodology comparing the results of \textsc{Rils}--\textsc{Rols} to 14 other approaches over all problem instances from the both ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz}. Note that an indicator random variable is involved per each run of the algorithms. Thus, the score that comes into the statistical observation is an indicator (0 or 1) indicating weather or not an algorithm was able to return the exact model per each run. More precisely, we involved 1300 results into our statistical evaluation for each of the algorithms. Note that  results of some of the 14 competitors are not known from \url{https://cavalab.org/srbench/results/} due to some unknown reasons (most likely algorithm is terminated running out of memory or due to the chosen  value of an internal parameter). For these runs, we simply count a 0 for the score.

Initially, Friedman’s test was separately executed for all competitor approaches.      In those cases in which the null hypothesis $H_0$ was rejected ($H_0$ states that there are no statistical differences between the obtained results of all competitors) pairwise comparisons are further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots. In each CD plot, the (15) competitor approaches are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is drawn.   Three (one for each level of noise) CD plots are shown in Figure~\ref{fig:cd-plots-exact-pcts} per each level of noise. The following conclusions may be drawn from there.

\begin{itemize}
	\item  Concerning the instance problems where no noise is included, \textsc{Rils}-\textsc{Rols} method returns the best average ranking in terms of the number of obtained exact models.  The second best average ranking is achieved by \textsc{AI-Feynman} algorithm. There is  statistically significant difference between these two approaches w.r.t.\ number of found exact models. The third best approach is \textsc{Apf-Fe},  which is far behind \textsc{Rils}-\textsc{Rols} w.r.t.\ delivered average ranking.  
	\item    Concerning the instance problems with the included  level of noise of 0.001, \textsc{Rils}-\textsc{Rols} method again obtains the best average ranking in terms of the found exact models.   The second best ranking is obtained by \textsc{AI-Feynman} algorithm. There is again a statistically significant difference between the results obtained by these two approaches. The third best approach is \textsc{Afp-Fe} which performs statistically worse than \textsc{AI-Feynman} approach.  
	\item  Concerning the instance problems with the included noise level of  0.01, again, \textsc{Rils-Rols} method produces the best average ranking according to positively evaluated runs, i.e. the runs with obtained exact models; the second best approach w.r.t.\ average ranking is \textsc{Afp-Fe}, followed by \textsc{Dsr}, and \textsc{Afp}. These latter three approaches perform statistically equal, but significantly worse than \text{Rils}-\textsc{Rols} method. 	The fourth best approach according to delivered average ranking is \textsc{AI-Feynman} which is on pair with \textsc{gplearn}. Note that the average ranking of \textsc{AI-Feynman} method declines by increasing  the level of noise in the input data. Thus, despite being efficient on no-noisy data, it seems that \textsc{AI-Feynman} is very sensitive to the level of noise increase, which is not that dramatically conspicuous with our \textsc{Rils}-\textsc{Rols} method.
